\chapter{Introduction}
\label{ch:introduction}

\section{Motivation}
\label{sec:intro-motivation}

Sign languages are the primary means of everyday communication for
tens of millions of deaf and hard-of-hearing people worldwide, and
they are linguistically as expressive and as fully developed as any
spoken language. They are nevertheless under-served by mainstream
consumer technology: voice assistants, captioning systems, and
translation tools all assume an acoustic input channel that is, by
definition, unavailable to the populations sign languages serve. The
resulting communication barrier between signing and non-signing
communities has not been closed even by the widespread availability of
camera-equipped smartphones.

Recent advances in computer vision, particularly on-device pose
estimation and compact deep-learning models, have made automatic sign
language recognition a plausible candidate for consumer accessibility
tools. Until recently this was a research topic accessible only on
workstation-class hardware. Several published systems now reach competitive
accuracy on standard benchmarks, but the literature has so far
emphasised accuracy over deployability; off-the-shelf systems with
state-of-the-art numbers rarely come with model-size or on-device
latency figures, and the gap between competitive on a server and
usable on a phone remains wide. The present thesis is motivated by
this gap.

\section{Problem Statement}
\label{sec:intro-problem}

This thesis addresses isolated American Sign Language recognition in
two complementary regimes that together cover most of the input an
everyday accessibility tool would have to handle:

\begin{enumerate}
  \item \textbf{Letter (fingerspelling) recognition:} given a single
        video frame of a static hand pose, classify it as one of the
        29 classes of the ASL Alphabet dataset (the 26 English
        letters together with the three auxiliary
        \emph{del}, \emph{space}, and \emph{nothing} classes). This
        task is dominated by static handshape and corresponds to the
        spelling out of proper nouns and loanwords.
  \item \textbf{Word-level recognition:} given a short video clip of
        a signer producing one isolated sign, classify it as one of
        the 100 most frequent signs of the Word-Level American Sign
        Language (WLASL) dataset. This task is dominated by motion,
        handshape transitions, and two-handed coordination.
\end{enumerate}

In both regimes the system must run end-to-end on a consumer Android
smartphone with no server-side inference. This deployability
constraint is treated not as an afterthought but as a primary design
axis: every architectural decision is evaluated against both
classification accuracy and the resulting model-size and latency
footprint.

\section{Scope and Contributions}
\label{sec:intro-scope}

The thesis is framed as a comparative deployability study rather than
as an accuracy contest. Its scope is deliberately bounded:

\begin{itemize}
  \item Only the WLASL-100 subset of the WLASL benchmark is used for
        word-level evaluation; the larger 300, 1{,}000, and 2{,}000
        class subsets are out of scope.
  \item Only American Sign Language is considered; other regional
        sign languages are out of scope.
  \item Only isolated signs are recognised; continuous sign language
        translation is out of scope.
  \item All inference is on-device; no cloud or server-side component
        is built.
\end{itemize}

The head-to-head image-vs-landmark comparison promised by the title
is realised in full at the letter task, where an EfficientNet-B0
image classifier and a landmark MLP are evaluated on the same
ASL Alphabet split. At the word task no image-based baseline is
included, because on-device image-based word recognition
(e.g.\ I3D, MViT, video-transformer variants) falls outside the
size and throughput budget that defines this thesis; the
comparative dimension at the word level is therefore between
single-seed and five-seed-ensemble variants of the landmark
BiLSTM rather than across architectural families.

Within this scope the contributions of the thesis are the following.

\begin{enumerate}
  \item An image-based fingerspelling recogniser built on an
        EfficientNet-B0 backbone, fine-tuned on the ASL Alphabet
        dataset and exported to a lightweight on-device inference
        runtime.
  \item A landmark-based fingerspelling recogniser built on a
        compact multilayer perceptron over MediaPipe hand keypoints,
        evaluated on the same ASL Alphabet split as the image-based
        recogniser so that the two letter-task pipelines can be
        compared directly.
  \item A landmark-based word-level recogniser, in which MediaPipe
        Hands landmarks are fed into a two-layer bidirectional LSTM,
        trained and evaluated on WLASL-100 under the dataset's
        published protocol.
  \item A WLASL-100 data-recovery pipeline that reconstructs
        a \num{1465}-clip working corpus from a partially link-rotted
        public source (a \SI{44}{\percent} increase over the
        \num{1013}-video Kaggle bundle alone) and a forward-additive
        ablation that quantifies the resulting gain alongside the
        extraction-pipeline and training-recipe contributions.
  \item A reproducible evaluation that reports Top-1 accuracy, model
        size, and on-device throughput side by side, and situates the
        resulting accuracy--footprint trade-off against published
        WLASL-100 baselines
        (Chapter~\ref{ch:test-results}).
\end{enumerate}

\section{Thesis Outline}
\label{sec:intro-outline}

The remainder of this document is organised as follows.
Chapter~\ref{ch:related-work} surveys prior work on image-based,
landmark-based, and temporal sign recognition, and on the mobile
deployment of such systems.
Chapter~\ref{ch:analysis-design} formalises the recognition problem
and presents the high-level design of the proposed two-track
pipeline.
Chapter~\ref{ch:implementation} describes how the design is realised
end-to-end, from data preparation through training to on-device
export. Chapter~\ref{ch:test-results} reports the quantitative
evaluation, including the WLASL-100 baseline comparison and the
on-device measurements. Chapter~\ref{ch:conclusion} summarises the
findings, discusses the limitations of the current system, and
outlines directions for future work.
