\chapter{Conclusion}
\label{ch:conclusion}

\section{Summary of Contributions}
\label{sec:conclusion-summary}

This thesis delivered an end-to-end American Sign Language
recognition system that runs entirely on a consumer Android phone.
It reports the evaluation metrics specified in the project proposal
(Top-1 and Top-5 accuracy, F1 score, confusion matrix, on-device FPS
and model footprint) and additionally clears the numeric acceptance
bars adopted for this thesis as concrete pass/fail targets on top
of those metrics. The contributions are summarised below in the
order in which they are developed in the preceding chapters.

\paragraph{Two complementary letter classifiers.}
Chapter~\ref{ch:implementation} described the design and training of
two ASL fingerspelling classifiers that occupy opposite ends of the
size and speed trade-off: an image-based EfficientNet-B0 and a
landmark-based MLP that consumes the 63-dimensional MediaPipe Hands
output and ships in a \SI{0.24}{MB} checkpoint, \(\sim\)65$\times$
smaller than the image network. On the held-out ASL Alphabet test
partition the two classifiers reach \SI{100}{\percent} and
\SI{99.38}{\percent} Top-1 accuracy respectively. Because the
dataset is single-contributor, uniform-background, and
fixed-distance, these offline figures are read in this thesis as a
sanity check that the training pipeline converges and that the
exported checkpoints preserve accuracy
(Section~\ref{sec:results-letters}); the actual capability signal
is sought in the on-device measurements that follow. The offline
numbers also provide the controlled benchmark against which the
architectural comparison is made. Because the
deployability of the system is the central objective of this
thesis, the offline comparison is paired with an on-device manual
evaluation
(Section~\ref{sec:results-qualitative-letters}) that adds a second
layer of evidence under realistic camera and inference conditions:
on the entry-level TECNO the image network drops to
\SI{92.9}{\percent} under the budget phone's camera pipeline while
the landmark MLP retains \SI{100}{\percent} on both test devices.
The landmark MLP is the correct deployment choice for the letter
task on budget hardware: it is more accurate than the image
network in the on-device manual test where the offline ranking
inverts, ships in a checkpoint roughly $65\times$ smaller, and
runs at lower per-frame latency on the same device
(Section~\ref{sec:results-on-device}). This is a converging
verdict on three independent axes, not a marginal preference. The
signer-disjoint generalisation of these classifiers, which the
random split of the public dataset does not assess, is discussed as
a limitation in Section~\ref{sec:conclusion-limitations}.

\paragraph{A word classifier that clears the adopted acceptance bar
with a hands-only feature set.}
Chapter~\ref{ch:implementation} described a five-seed bidirectional
LSTM trained on a two-handed 128-dimensional landmark feature stream.
The word classifier clears the adopted hard acceptance bar. The
five-seed logit-averaging ensemble reaches \SI{56.22}{\percent}
Top-1 and \SI{81.09}{\percent} Top-5 on the WLASL-100 test
partition under its canonical person-mixed split
(Table~\ref{tab:results-words-seeds}), passing the
\SI{40}{\percent} Top-1 acceptance threshold by 16.22 percentage
points; every individual seed in the pool also clears the bar, and
the paired-bootstrap gain over the per-seed mean of
\SI{52.14}{\percent} is significant at $p = 0.015$. The role of
the ensemble at this number is a variance-reduction one and is
discussed in Section~\ref{sec:results-words}; the person-mixed
protocol places the offline figure as a literature-comparable
benchmark point rather than a signer-independent generalization
estimate (Section~\ref{sec:conclusion-limitations}). Within the
landmark-only family, the position is owned explicitly: hands-only
landmark inference lands above the 2020 Pose-TGCN baseline
(\SI{55.43}{\percent}) but below the contemporaneous hands-only
PBRC pipeline of Singh et al.~\cite{singh2026pbrc}
(\SI{60.85}{\percent}, ${-}\SI{4.63}{}$ points) and the
pose-transformer SPOTER (\SI{63.18}{\percent},
${-}\SI{6.96}{}$ points). The thesis-level word-task
contribution is therefore not formulated as an accuracy claim. It
is the joint same-protocol report: this work is the first
landmark-only WLASL-100 study to publish on-device latency and a
megabyte-scale model footprint alongside offline accuracy under
the canonical protocol
(Sections~\ref{sec:results-on-device}--\ref{sec:results-qualitative}).

\paragraph{A forward-additive ablation that decomposes the gain from
\SI{20}{\percent} to \SI{60}{\percent} into three methodological pillars.}
The final word classifier was not produced by a single design choice but
by a nine-step trajectory documented in
Table~\ref{tab:results-ablation}, in which each modification was
isolated by retraining three independently seeded models on the same
WLASL-100 split. The 40-point gain decomposes cleanly into three
contributions of comparable scientific weight. \emph{Data engineering}
contributes +16.08 validation points: the yt-dlp-based recovery of 448
link-rotted videos (\num{1013}\,$\to$\,\num{1465}) is the largest single
step in the entire table and surfaces dataset-size sensitivity as the
dominant lever on this WLASL-100 setup. The recovery itself is
WLASL-specific (it repairs this benchmark's link-rotted YouTube
manifest rather than introducing a generalisable technique), but the
finding -- that the recurrent classifier sits at chance plus a
memorization band under \num{1013} videos and clears 36\,\% with
\num{1465} -- is the operative takeaway and applies to any future
study that approaches WLASL-100 from a comparable training budget. The \emph{extraction pipeline}, which adds
two-hand landmarks with presence flags and WLASL frame-window cropping,
contributes a further +10.29 points without changing the training
recipe. The \emph{training-recipe refinements} (light sequence
augmentation, stronger frame dropout, time-scale, mixup, the long-tail
\texttt{WeightedRandomSampler}, and the wider hidden state) together
add the remaining ${\sim}14$ points. The ablation also surfaces a
methodological detail worth reporting: the wider hidden state is mildly
negative in isolation ($-1.24$ points) but contributes a $+4.39$-point
\emph{interaction effect} once the full regularization stack is in
place (Section~\ref{sec:results-ablation-interaction}), and removing
it from the full final configuration costs $-6.47$ test points; its
inclusion in the deployed model is therefore justified by joint
adoption rather than by a capacity argument alone. The five-seed ensemble reaches
\SI{60.08}{\percent} on the validation set and \SI{56.22}{\percent} on
the test set, lifting the five-seed test mean of \SI{52.14}{\percent}
by a paired-bootstrap significant +4.08 points ($p = 0.015$) while
remaining statistically indistinguishable from the best single seed on
the same 201-clip partition (McNemar $p = 1.000$). Together, the table
and the interaction observation are reusable as a recipe for related
landmark-based sequence classification problems on small datasets.

\paragraph{An on-device Android application that exposes all four
classifiers.}
Chapter~\ref{ch:implementation} described a CameraX-driven Android
application that runs the four classifiers through TensorFlow Lite
with the XNNPACK CPU delegate, displays a Top-3 prediction panel and
a self-reported FPS/latency badge, and recovers from MediaPipe's
cold-start timestamp anomaly through a volatile guard around
\texttt{detectAsync}. The Words tab implements the single-shot lock
state machine described in Section~\ref{sec:impl-android-pipeline}:
a \(\sim\)1~s pre-collection countdown, single inference after the
buffer fills, on-screen latching of the result, and automatic
release on an eight-frame hand-down signal. No cloud inference, no
network connectivity, and no server-side component is required at
inference time.

\paragraph{An eight-cell on-device performance characterisation.}
Section~\ref{sec:results-on-device} reported FPS and latency for
each of the four classifiers on two phones: the entry-level
TECNO SPARK 20C (Helio G36) and the flagship Xiaomi 15T Pro
(Dimensity 9400+). The headline result is that, after the
GPU delegate was removed for cold-start stability, the cost of
CPU-only execution is invisible on the flagship (Ensemble at
\SI{30}{FPS} / \SI{80}{ms}) and prohibitive on the entry-level
device (Ensemble at \SI{17}{FPS} / \SI{665}{ms}). The matrix
produces a concrete deployment guideline: on entry-level
hardware only the landmark MLP is interactive; on flagship
hardware all four classifiers are usable and model selection
reduces to the offline accuracy / size trade-off.

\paragraph{A two-device single-user manual evaluation that bounds
the offline {$\rightarrow$} on-device gap.}
Section~\ref{sec:results-qualitative} reported a one-trial-per-class
manual test of the letter classifiers on both devices (29 trials per
cell, 28 measurable) and of the word classifiers on the Xiaomi (100
trials per model). The single-user OOD gap relative to the offline
test partitions was at most seven percentage points for the letter
classifiers and \SIrange{7}{14}{\percent} points for the word
classifiers, with both word classifiers clearing the manual-test
acceptance bars adopted in Section~\ref{sec:results-words-acceptance}
(\SI{38.0}{\percent} single, \SI{49.0}{\percent} ensemble) and a
structured set of attractor classes
(\texttt{white}, \texttt{walk}, \texttt{drink}, \texttt{full})
identified as the dominant failure mode. As a single-user
feasibility probe, the manual test is consistent with the
deployability claim that motivates the thesis; it is not, on its own,
a controlled measurement of classifier accuracy under deployment
conditions.

\section{What Was Not Done}
\label{sec:conclusion-not-done}

A clear statement of what falls outside the scope of the present
work is as important as a statement of what was achieved, both for
honest reporting and because the senior-project guidelines require
it. The boundaries listed here are deliberate; each was fixed at
the proposal stage in
Chapter~\ref{ch:introduction} and observed throughout the
implementation and evaluation chapters.

\begin{itemize}
  \item \textbf{Larger WLASL subsets were not evaluated.}
        Word-level training and evaluation are restricted to the
        WLASL-100 subset. The larger 300, 1{,}000, and 2{,}000-class
        subsets of the WLASL benchmark are not addressed; the
        person-mixed test protocol, the bidirectional LSTM
        backbone, and the on-device export pipeline are evaluated
        only at the 100-class operating point.
  \item \textbf{Sign languages other than ASL were not considered.}
        Only American Sign Language is recognised. No data, no
        labels, and no model from any other regional sign language
        (e.g.\ T\"urk \.I\c{s}aret Dili, British Sign Language,
        Deutsche Geb\"ardensprache) enters the pipeline. The
        landmark-based front end is in principle language-agnostic,
        but cross-language transfer is not investigated.
  \item \textbf{Continuous sign language translation was not
        attempted.} Both pipelines classify a single isolated
        unit, either a static letter or a pre-segmented word clip,
        and produce a class label rather than a gloss sequence or a
        translation into written language. Co-articulation between
        adjacent signs, automatic temporal segmentation of a
        free-flowing signing stream, and the mapping from gloss to
        natural-language sentences are out of scope.
  \item \textbf{No cloud or server-side inference component was
        built.} All recognition is performed on the consumer
        Android device, using the exported TensorFlow Lite
        artefacts. The system does not call any remote API at
        inference time and does not require network connectivity.
        Server-side training, ensembling, or re-ranking, all
        common in commercial accessibility tools, is therefore not
        part of the deliverable.
  \item \textbf{Signer-level analysis was not performed.}
        The ASL Alphabet dataset does not expose signer identity in
        its public release, and the WLASL-100 evaluation follows
        the dataset's published person-mixed split as a single
        fixed partition. Per-signer accuracy breakdowns,
        cross-signer adaptation, and few-shot personalisation to a
        new signer are not reported.
  \item \textbf{User-study evaluation with deaf or
        hard-of-hearing participants was not conducted.} The
        present evaluation is purely technical: classification
        accuracy on held-out data and on-device throughput on the
        target hardware. A field study with members of the
        intended user community would be the appropriate next step
        before any deployment claim, but it was not within the
        scope of a one-semester senior-project thesis and is left
        to future work.
\end{itemize}

\section{Limitations}
\label{sec:conclusion-limitations}

The previous section listed what was placed outside the scope of
the thesis by design. This section lists the residual limitations
of \emph{what was actually built}: properties of the delivered
system that constrain how its measurements should be interpreted
and where its applicability ends. They are grouped into
architectural limitations, which follow from design choices,
data-side limitations, which follow from the public datasets the
system was trained on, and performance-side limitations, which
follow from the on-device measurements of
Section~\ref{sec:results-on-device} and the manual evaluation of
Section~\ref{sec:results-qualitative}.

\paragraph{Hands-only landmark stream.}
Both landmark-based classifiers consume the 21~hand-landmark
output of MediaPipe Hands and nothing else. ASL signs frequently
carry meaning in components that this representation discards:
body location (signs anchored to the chest, shoulder, or face),
upper-arm motion (which a two-handed sign uses to coordinate its
hands), and non-manual markers such as eyebrow position or mouth
shape. The error analysis of
Section~\ref{sec:results-errors} attributes the most frequent
WLASL-100 confusion pairs precisely to this missing context: the
five top pairs all share handshape and gross hand motion and
differ only in body location or facial marker. A model that adds
body pose, and especially facial landmarks, would resolve a
fraction of these errors at near-zero inference cost on the
target hardware (the corresponding extension is the first item of
Section~\ref{sec:conclusion-future}).

\paragraph{Single-hand letter pipeline.}
The landmark-based fingerspelling MLP is trained on the dominant
hand only; if MediaPipe detects two hands in the frame, only the
first is consumed and the rest is discarded. This is appropriate
for the bulk of the ASL Alphabet, since most letters use a single
hand, but it leaves regional two-hand variants and the auxiliary
classes under-supported. The image-based EfficientNet
classifier does not have this limitation, since it operates on
raw pixels, but it correspondingly cannot benefit from the
explicit two-handed structure that the LSTM word model uses.

\paragraph{MediaPipe detection floor.}
Both landmark pipelines are gated on MediaPipe's hand detector.
On the ASL Alphabet test split, MediaPipe failed to recover a
hand on \SI{26.91}{\percent} of test images
(Section~\ref{sec:impl-mlp}), including the entire
\texttt{nothing} class and a thumb-tucked subset of \textit{M}
and~\textit{N}. The Android demo handles these frames gracefully
by greying out the landmark-based output, but the practical
implication is the same: under conditions where MediaPipe fails
(low-resolution input, motion blur, partial occlusion, unusual
lighting), the landmark pipelines simply do not produce a
prediction. The image-based pipeline is the appropriate fallback
on such frames; the system does not currently switch between the
two automatically.

\paragraph{ASL Alphabet generalisation cannot be inferred from
the reported test accuracy.}
The \SI{100}{\percent} EfficientNet and \SI{99.38}{\percent} MLP
numbers reported in Section~\ref{sec:results-letters} are
upper-bound figures, obtained on a random 70/15/15 split of a
single-source Kaggle dataset whose images were collected by one
contributor in a controlled indoor setting. The dataset does not
expose signer identity in its public release, so a
signer-disjoint split is not constructible; the random split
allows the same person, lighting, and background to appear in
both partitions. The reported accuracy therefore measures
in-distribution discrimination, not generalisation to a new
signer in a new room with a new camera. The real-device
measurements of Section~\ref{sec:results-qualitative} are the
appropriate ground truth for the latter question.

\paragraph{WLASL-100 person-mixed split and the unmeasured
generalization band.}
This is the headline structural limitation of the present
evaluation and is stated as such. The WLASL-100 evaluation uses
the dataset's published canonical split, which is person-mixed:
the same signer can appear in both the train and the test
partitions. This is a known property of the WLASL benchmark and
is shared by all published baselines in
Table~\ref{tab:results-wlasl-lit}, so the relative comparisons in
that table are on equal terms. The absolute number, however, is
inflated by construction relative to a stricter signer-disjoint
protocol, and a system that reaches \SI{56.22}{\percent} on the
mixed split should not be assumed to reach the same number on
data from a signer that the training set has never seen.

Two things were therefore measured cleanly by the word-task
evaluation of Chapter~\ref{ch:test-results}, and one was not.
What \emph{was} measured cleanly is, first, a
protocol-comparable benchmark point against published WLASL-100
methods on the canonical person-mixed split
(Section~\ref{sec:results-words-literature}); and second, the
on-device throughput, latency, and known-sign single-user
deployability of all four classifiers on two distinct Android
devices (Sections~\ref{sec:results-on-device} and
\ref{sec:results-qualitative}). What was \emph{not} measured is
the signer-independent generalization band of the word
classifier: the offline figure is an inflated-by-design upper
bound on the canonical protocol, and the on-device manual test
is a known-sign single-user upper bound under naturalistic
camera conditions (the explicit familiarisation step of
Section~\ref{sec:results-qualitative-protocol} makes this an
upper bound rather than a representative production-conditions
sample). The actual signer-independent accuracy lies somewhere
in the band these two figures bracket, and the present thesis
does not narrow that band beyond bracketing it. A
signer-disjoint re-evaluation on a future release of the WLASL
metadata, paired with the multi-signer field study listed in
Section~\ref{sec:conclusion-future}, is the appropriate
follow-up.

\paragraph{Statistical power of the 201-clip word test set.}
The WLASL-100 canonical test partition contains only 201 clips, so a
single misclassification moves Top-1 by \SI{0.50}{percentage point}
and the 95\,\% Wilson binomial confidence interval around any Top-1
estimate spans roughly \(\pm 7\) points. The per-seed CIs in
Table~\ref{tab:results-words-seeds} all overlap heavily and the
ensemble's offline Top-1 is statistically indistinguishable from the
best single seed on this partition (McNemar $p = 1.000$,
Section~\ref{sec:results-words}). The ensemble's offline contribution
is therefore a variance-reduction one (removing the seed-selection
gamble) rather than a tight absolute-accuracy claim; the on-device
manual test of Section~\ref{sec:results-qualitative-words} (\(n=100\))
is the second statistical signal that the present evaluation carries.
A larger held-out partition would tighten both readings and is
listed as future work in Section~\ref{sec:conclusion-future}.

\paragraph{Isolated-only word recognition.}
The word classifier expects a temporally pre-segmented clip
containing exactly one sign. It does not detect when a sign
starts or ends in a continuous signing stream, and it does not
model co-articulation between adjacent signs. The Android demo
applies the same fixed 32-frame window logic at inference time
and therefore inherits this assumption: the user must produce
each word in isolation for the displayed classification to be
meaningful. A real accessibility tool would have to add a
segmentation stage in front of the classifier, an extension that
is identified as a high-impact item in
Section~\ref{sec:conclusion-future}.

\paragraph{Asymmetric latency cost of CPU-only execution on
entry-level hardware.}
To eliminate a cold-start \texttt{MediaPipeException} in the
\texttt{LIVE\_STREAM} path of MediaPipe Hands, the deployed Android
build runs all four classifiers on the CPU through XNNPACK and does
not enable the TensorFlow Lite GPU delegate
(Section~\ref{sec:results-on-device-cpu}). The cost of this decision
is invisible on flagship hardware: on the Xiaomi 15T Pro all four
classifiers ran at \SI{30}{FPS} with end-to-end latencies of at most
\SI{90}{ms}. On entry-level hardware, however, the same choice is
prohibitive. On the
TECNO SPARK 20C, EfficientNet-B0 and the BiLSTM single model both
exceeded \SI{200}{ms} mean latency and the five-seed ensemble
reached \SI{665}{ms} (Table~\ref{tab:on-device-words}), at which
point the application is no longer interactive. The trade-off taken
for build stability is therefore not uniform across the deployment
spectrum: it is free on flagship phones and disqualifying for the
ensemble on budget phones. A future build that exposes a per-device
delegate choice (GPU on devices where the cold-start path is stable,
CPU elsewhere) would partly recover this cost, but the present
thesis does not include that work.

\paragraph{Camera-pipeline sensitivity of the image-based letter
classifier.}
The manual test
(Section~\ref{sec:results-qualitative-letters}) revealed a behavior
that the offline test partition did not. EfficientNet-B0, which
reaches \SI{100}{\percent} on the offline letter test set, drops to
\SI{92.9}{\percent} on the entry-level TECNO and stays at
\SI{100}{\percent} on the Xiaomi. The two errors on the TECNO
(\textit{D}~$\rightarrow$~\textit{K/R} and
\textit{K}~$\rightarrow$~\textit{W}) involve hand shapes that the
landmark MLP classified correctly on the same device with full
confidence; their probable cause is the budget phone's image-signal
processing pipeline (sharper highlights, harsher white-balance
correction, slight motion blur) interacting with a pixel-based
backbone that was trained on the Kaggle ASL Alphabet dataset's
controlled-indoor imagery. The landmark abstraction discards
precisely this kind of camera-conditioned variation. The practical
consequence is that the image classifier's offline-vs-deployed
accuracy gap is itself device-conditioned, and a deployment that
guarantees image-network accuracy across the budget tier would
require either a camera-aware retraining set or a runtime fallback
to the landmark variant.

\paragraph{Single-user OOD ceiling on the manual evaluation.}
The manual evaluation of Section~\ref{sec:results-qualitative} was
conducted by a single human signer (the author). One signer cannot
sample the population of hand morphologies, signing styles,
fingerspelling speeds and ambient camera conditions that a real
deployment would encounter, so the manual-test accuracy numbers are
not representative samples of production accuracy in either
direction. A second qualification applies to the Words manual test
specifically: because the WLASL-100 vocabulary is not part of the
author's active sign repertoire, each Words trial was preceded by a
short familiarisation step against the WLASL reference clip, and
the recorded gesture was the author's best-effort production of
that reference (Section~\ref{sec:results-qualitative-words}). The
resulting Words manual-test accuracy therefore measures the
classifier under a known-sign assumption and constitutes an upper
bound on the production accuracy that a first-time user without
this familiarisation would obtain. They are useful for two narrower
purposes: verifying that the proposal's qualitative criterion
(``the recognised letter or word appears on screen'') is satisfied
in real use, and bounding the offline-to-on-device gap on a single
OOD draw under that known-sign assumption. A
representative production accuracy number would require a
multi-signer evaluation that the present scope did not include,
and is the user-study item listed in
Section~\ref{sec:conclusion-not-done} and revisited as the highest-impact
follow-up in Section~\ref{sec:conclusion-future}.

\section{Future Work}
\label{sec:conclusion-future}

The scope boundaries and limitations identified above suggest a
number of concrete extensions that would be natural follow-ups to
the present thesis. They are listed in increasing order of effort.

\paragraph{Augmenting the landmark front end with body pose.}
The current word-level pipeline consumes only the 21~hand landmarks
of MediaPipe Hands, dropped to zero when the hand leaves the frame.
WLASL signs, however, frequently involve the upper body: signs are
located in space relative to the chest, the dominant hand often
references the non-dominant arm, and two-handed signs depend on
coordinated motion of both hands. Combining MediaPipe Hands with
the body-pose joints of MediaPipe Pose, in the same spirit as
Pose-TGCN and the pose-guided RGB fusion of
Hosain et~al.~\cite{hosain2021fusion} at the body level, would add
this missing context at a near-zero inference cost on the mobile
target, because both extractors
already run on-device. A direct comparison between the present
hands-only landmark stream and a Pose+Hands stream, with the rest
of the architecture held fixed, is the most immediate extension.

\paragraph{Replacing the recurrent temporal model with a
transformer encoder.}
Section~\ref{sec:related-temporal} surveyed the move from recurrent
to transformer-based temporal models on WLASL, with the Sign
Pose-Based Transformer of Boh\'a\v{c}ek and
Hr\'uz~\cite{bohacek2022spoter} reaching 63.18\% Top-1 on the
100-class subset using pose-only inputs. A natural follow-up to the
present bidirectional LSTM is to replace its recurrent backbone
with a comparably sized pose transformer encoder, training and
evaluating it under the same person-mixed protocol and the same
on-device export pipeline. The interesting question is not whether
a transformer can match an LSTM on this dataset, since the prior
literature suggests it can. The real question is whether the
resulting model remains within the mobile size and latency budget
that motivates this thesis.

\paragraph{Two-handed handling on the letter task.}
The landmark-based fingerspelling MLP is trained on a single hand
and falls back to a zero-vector when no hand is detected. Letters
that involve two hands in some regional variants, and several of
the auxiliary classes, would benefit from a model that consumes
both hands when both are present. This is a small architectural
change, essentially doubling the input dimension and zero-padding
the secondary hand, but it would require re-training on a dataset
that preserves both-hand annotations.

\paragraph{Real-time word boundary detection.}
The word-level pipeline is evaluated under the WLASL-100 protocol,
which assumes pre-segmented isolated clips. A usable accessibility
tool would have to detect the start and end of each sign in a
continuous signing stream. A lightweight on-device boundary
detector, for example a binary classifier over short windows of the
same landmark stream or an energy-based segmenter over hand
velocity, could be inserted in front of the existing word
classifier without changing the latter. This extension is mostly
engineering rather than research, but it is the single change with
the largest impact on perceived usability.

\paragraph{Cross-lingual transfer.}
Because the landmark front end is independent of the spoken
language of the signer, the same MLP and LSTM architectures could
in principle be re-trained on a Turkish Sign Language
fingerspelling and isolated-sign dataset. Investigating how much
of the WLASL-100 trained weight can be reused as a warm start,
versus how much must be re-learned from scratch, would address a
practical question for Turkish-speaking deaf communities and would
be a natural continuation of this thesis in a graduate-level
follow-up.

\paragraph{User study with the intended community.}
Finally, all of the above extensions are technical. The single
extension that would have the highest impact on whether the system
is actually useful is a field evaluation with deaf and
hard-of-hearing participants. Such a study would measure task success rather than
classification accuracy. The relevant questions are whether the
participant can fingerspell a name and have the application recognise
it, and whether a hearing interlocutor can follow a 100-word
vocabulary exchange. A study at this level would also produce design
feedback (camera framing prompts, confidence display, error-recovery
affordances) that no offline benchmark can provide.
